Optimizing Abstract Abstract Machines

### 2012-11-14 paper/paper.tex
Machine components, work on initial analyzer.

The \emph{abstracting abstract machines} (AAM)
approach~\cite{dvanhorn:VanHorn2011Abstracting,dvanhorn:VanHorn2012Systematic}
to deriving program analyses provides a systematic way of transforming a
programming language semantics in the form of an abstract machine into a family
of abstract interpreters.
%
The approach parameterizes these families with policies for regulating analytic
precision.
%
While flexible and robust, the AAM approach unfortunately yields analyzers with
poor performance relative to hand-optimized analyzers.
%
Our work takes aim squarely at this ``efficiency gap,'' and narrows it in an
equally systematic way.

%
% Specifically, these policies tune precision for control,
% the environment, the store, and base value domains.

By taking a
machine-oriented view of computation, AAM makes it possible to design,
verify, and implement program analyzers for realistic language
features typically considered difficult to model.  The approach was
originally applied to features such as higher-order functions,
stack inspection, exceptions, laziness, first-class continuations, and
garbage collection.  It has since been used to verify actor-
\cite{local:DOsualdo:12A} and
thread-based~\cite{dvanhorn:Might2011Family} parallelism and
behavioral contracts~\cite{dvanhorn:TobinHochstadt2012Higherorder}; it
has been used to model Coq~\cite{local:harvard},
Dalvik~\cite{local:dalvik}, Erlang~\cite{local:DOsualdo:12B},
JavaScript~\cite{local:DBLP:journals/corr/abs-1109-4467}, and
Racket~\cite{dvanhorn:TobinHochstadt2012Higherorder}.

@@

%% Using this approach has also lead to new insights in pushdown
%% abstractions of programs capable of precisely modeling the program
%% stack, even in the presence of control effects.

%% \subsection{Notation and prerequisites}

%% Prereqs: Semantics Engineering \cite{dvanhorn:Felleisen2009Semantics},
%% AAM \cite{dvanhorn:VanHorn2011Abstracting}
%% \cite{dvanhorn:VanHorn2012Systematic}.

%% Notation: concrete examples of programs under analysis is given in
%% (monochrome) Scheme notation.  Code is given in (syntax colored)
%% Racket.

@@

At each step during the initial development, we evaluate the
implementation on a benchmark from Vardoulakis and
Shivers~\cite{dvanhorn:Vardoulakis2011CFA2} that tests distributivity
of multiplication over addition on Church numerals.
%% \footnote{It is
%%   written here using Scheme notation and makes use of the shorthands:
%% %
%% \begin{alltt}
%%    (define (((f x) y) z) ...) \(\equiv\)
%%
%%    (define f (\(\lambda\) (x) (\(\lambda\) (y) (\(\lambda\) (z) ...))))
%% \end{alltt}
%%
%% \begin{alltt}
%%    \church{n} \(\equiv\) (\(\lambda\) (s) (\(\lambda\) (z) (s\(\subscript1\) (s\(\subscript2\) ... (s\(\subscript{n}\) z)))))
%% \end{alltt}}
The example, while small, is informative:
\begin{enumerate}
\item it can be written in most modern programming languages,
%
\item it was designed to stress an analyzer's ability to deal with
  complicated environment and control structure arising from the use
  of higher-order functions to encode arithmetic, and
%
\item it proves to be the \emph{least} improved benchmark of the
  complete suite considered in section~\ref{sec:eval}, and thus it
  serves as a good sanity check and lower-bound for each of the
  optimization techniques considered.
\end{enumerate}

@@

%% \newcommand{\church}[1]{\(\ulcorner{\tt #1}\urcorner\)}
%% \begin{figure}
%% \hrule
%% \vspace{2mm}
%% \begin{alltt}
%% (define (zero? n) ((n (\(\lambda\) (zx) #f)) #t))
%% (define (((mult m1) m2) mf) (m2 (m1 mf)))
%% (define ((((plus p1) p2) pf) x)
%%   ((p1 pf) ((p2 pf) x)))

%% (define (((pred n) rf) rx)
%%   (((n (\(\lambda\) (g) (\(\lambda\) (h) (h (g rf)))))
%%     (\(\lambda\) (ignored) rx))
%%    (\(\lambda\) (id) id)))

%% (define ((church=? e1) e2)
%%   (if (zero? e1) (zero? e2)
%%       (if (zero? e2) #f
%%           ((church=? (pred e1)) (pred e2)))))

%% ;; multiplication distributes over addition
%% ((church=? ((mult \church2) ((plus \church1) \church3)))
%%  ((plus ((mult \church2) \church1)) ((mult \church2) \church3)))
%% \end{alltt}
%% \caption{Church-numeral calculation}
%% \label{fig:church}
%% \end{figure}

\begin{figure}[t]
%% \hrule
%% \vspace{2mm}
%% \begin{alltt}
%% (define (id x) x)
%% (define (f n)
%%   (if (<= n 1) 1 (* n (f (- n 1)))))
%% (define (g n)
%%   (if (<= n 1) 1 (+ (* n n) (g (- n 1)))))
%% (print (+ ((id f) 3) ((id g) 4)))
%% \end{alltt}
%% \hrule
\begin{center}
\begin{tabular}{ccc}
\raisebox{1ex-\height}{
\includegraphics[height=3.5in]{introspective-base.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=3.5in]{introspective-lazy.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=3in]{introspective-lazyc.pdf}}
\\
(a) Baseline
&
(b) Lazy
&
(c) Compiled (\& lazy)
\end{tabular}
\end{center}
\caption{Example state graphs for the program above.  Part (a) shows
  the result of the baseline analyzer.  It has long ``corridor''
  transitions and ``diamond'' subgraphs that fan-out from
  nondeterminism and fan-in from joins.  Part (b) shows the result of
  performing nondeterminism lazily and thus avoids many of the diamond
  subgraphs.  Part (c) shows the result of abstract compilation that
  removes intepretive overhead in the form of intermediate states,
  thus minimizing the corridor transitions.  The end result is a more
  compact abstraction of the program that can be generated faster.}
\label{fig:state-graphs}
\end{figure}

@@

%\newpage
\section{An abstract$^{\text 2}$ machine for ISWIM}
\label{sec:aam}

ISWIM is a family of programming languages parameterized by a set of
base values and operations.  To make things concrete, we consider a
member of the ISWIM family with integers, booleans, and a few
operations.

\begin{figure}
\[
\begin{array}{l@{\qquad}rcl}
\text{Expressions} & \mathit{e} &=& \svar[^\mlab]\mvar\\
&&|& \slit[^\mlab]\mlit\\
&&|& \slam[^\mlab]\mvar\mexp\\
&&|& \sapp[^\mlab]\mexp\mexp \\
&&|& \sif[^\mlab]\mexp\mexp\mexp \\
\text{Variables}&\mvar &=& \syntax{x}\ |\ \syntax{y}\ |\ \dots\\
\text{Literals}&\mlit &=& \mnum\ |\ \mbln\ |\ \mop\\
\text{Integers}&\mnum &=& \syntax{0}\ |\ \syntax{1}\ |\ \syntax{-1}\ |\ \dots\\
\text{Booleans}&\mbln &=& \strue\ |\ \sfalse\\
\text{Operations}&\mop &=& \syntax{zero?}\ |\ \syntax{add1}\ |\ \syntax{sub1}\ |\ \dots
\end{array}
\]
\caption{Syntax of ISWIM}
\label{fig:syntax}
\end{figure}

Figure~\ref{fig:syntax} defines the (abstract) syntax of ISWIM.
Figure~\ref{fig:aam} defines the semantics of ISWIM as a machine
model.  Evaluation is defined as the set of states reachable by the
reflexive, transitive closure of the machine transition relation.  The
machine is a very slight variation on a standard abstract machine for
ISWIM in ``eval, continue, apply''
form~\cite{dvanhorn:Danvy:DSc}.  It can be
systematically derived from a definitional interpreter through a
continuation-passing style transformation and defunctionalization, or
from a structural operational semantics using the refocusing
construction of Danvy and
Nielsen~\cite{dvanhorn:Danvy-Nielsen:RS-04-26}.

>>>>

The \emph{abstracting abstract machines} (AAM)
approach~\cite{dvanhorn:VanHorn2011Abstracting,dvanhorn:VanHorn2012Systematic}
to deriving program analyses provides a straightforward way of
transforming a programming language semantics in the form of an
abstract machine, into a family of abstract interpreters parameterized
over policies for regulating the analytic precision of the control,
environment, store, and base value domains.  By taking a
machine-oriented view of computation, it becomes possible to design,
verify, and implement program analyzers for realistic language
features typically considered difficult to model.  The approach was
originally applied to features such as higher-order functions,
stack inspection, exceptions, laziness, first-class continuations, and
garbage collection.  It has since been used to verify actor-
\cite{local:DOsualdo:12A} and
thread-based~\cite{dvanhorn:Might2011Family} parallelism and
behavioral contracts~\cite{dvanhorn:TobinHochstadt2012Higherorder}; it
has been used to model Coq~\cite{local:harvard},
Dalvik~\cite{local:dalvik}, Erlang~\cite{local:DOsualdo:12B},
JavaScript~\cite{local:DBLP:journals/corr/abs-1109-4467}, and
Racket~\cite{dvanhorn:TobinHochstadt2012Higherorder}.

@@

@@

At each step during the initial development, we evaluate the
implementation on a benchmark from Vardoulakis and
Shivers~\cite{dvanhorn:Vardoulakis2011CFA2} that tests distributivity
of multiplication over addition on Church numerals.
The example, while small, is informative:
\begin{enumerate}
\item it can be written in most modern programming languages,
%
\item it was designed to stress an analyzer's ability to deal with
  complicated environment and control structure arising from the use
  of higher-order functions to encode arithmetic, and
%
\item it proves to be the \emph{least} improved benchmark of the
  complete suite considered in section~\ref{sec:eval}, and thus it
  serves as a good sanity check and lower-bound for each of the
  optimization techniques considered.
\end{enumerate}

@@

\begin{figure}[t]
\begin{center}
\begin{tabular}{ccc}
\raisebox{1ex-\height}{
\includegraphics[height=3.5in]{introspective-base.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=3.5in]{introspective-lazy.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=3in]{introspective-lazyc.pdf}}
\\
(a) Baseline
&
(b) Lazy
&
(c) Compiled (\& lazy)
\end{tabular}
\end{center}
\caption{Example state graphs for the program above.  Part (a) shows
  the result of the baseline analyzer.  It has long ``corridor''
  transitions and ``diamond'' subgraphs that fan-out from
  nondeterminism and fan-in from joins.  Part (b) shows the result of
  performing nondeterminism lazily and thus avoids many of the diamond
  subgraphs.  Part (c) shows the result of abstract compilation that
  removes intepretive overhead in the form of intermediate states,
  thus minimizing the corridor transitions.  The end result is a more
  compact abstraction of the program that can be generated faster.}
\label{fig:state-graphs}
\end{figure}

@@

%\newpage
\section{An analyzer for ISWIM}
\label{sec:aam}

In this section, we give a brief review of the AAM approach by
defining a sound analytic framework for a core higher-order functional
language: Landin's ISWIM~\cite{dvanhorn:Landin1966Next}.  In the
subsequent sections, we will explore optimizations for the analyzer in
this simplified setting, but scaling these techniques to realistic
languages is straightforward and has been done for the analyzer
evaluated in section~\ref{sec:eval}.

ISWIM is a family of programming languages parameterized by a set of
base values and operations.  To make things concrete, we consider a
member of the ISWIM family with integers, booleans, and a few
operations.
%
Figure~\ref{fig:syntax} defines the (abstract) syntax of ISWIM.  It
includes variables, literals (either integers, booleans, or
operations), $\lambda$-expressions for defining procedures, procedure
applications, and conditionals.  Expressions carry a label, $\mlab$,
which is drawn from an unspecified set and denotes the source location
of the expression; labels are used to disambiguate distinct, but
syntactically identical pieces of syntax.  We omit the label
annotation in contexts where it is irrelevant.

\begin{figure}
\[
\begin{array}{l@{\qquad}rcl}
\text{Expressions} & \mexp &=& \svar[^\mlab]\mvar\\
&&|& \slit[^\mlab]\mlit\\
&&|& \slam[^\mlab]\mvar\mexp\\
&&|& \sapp[^\mlab]\mexp\mexp \\
&&|& \sif[^\mlab]\mexp\mexp\mexp \\
\text{Variables}&\mvar &=& \syntax{x}\ |\ \syntax{y}\ |\ \dots\\
\text{Literals}&\mlit &=& \mnum\ |\ \mbln\ |\ \mop\\
\text{Integers}&\mnum &=& \syntax{0}\ |\ \syntax{1}\ |\ \syntax{-1}\ |\ \dots\\
\text{Booleans}&\mbln &=& \strue\ |\ \sfalse\\
\text{Operations}&\mop &=& \syntax{zero?}\ |\ \syntax{add1}\ |\ \syntax{sub1}\ |\ \dots
\end{array}
\]
\caption{Syntax of ISWIM}
\label{fig:syntax}
\end{figure}

\begin{figure}
\[
\begin{array}{l@{\qquad}rcl}
\text{Values} & \mval   &=& \clos{\mvar,\mexp,\menv}\ |\ \mlit\\
\text{States} & \mstate &=& \ev{\mexp,\menv,\msto,\mkont}\\
                       &&|& \co{\mkont,\mval,\msto}\\
                       &&|& \ap{\mval,\mval,\msto,\mkont}\\
\text{Continuations} & \mkont &=& \kmt\\
&&|& \kfn{\mval,\maddr}\\
&&|& \kar{\mexp,\menv,\maddr}\\
&&|& \kif{\mexp,\mexp,\menv,\maddr}
\end{array}
\]
\caption{Abstract machine components}
\end{figure}

Figure~\ref{fig:aam} defines the semantics of ISWIM as a machine
model.  Evaluation is defined as the set of states reachable by the
reflexive, transitive closure of the machine transition relation.  The
machine is a very slight variation on a standard abstract machine for
ISWIM in ``eval, continue, apply''
form~\cite{dvanhorn:Danvy:DSc}.  It can be
systematically derived from a definitional interpreter through a
continuation-passing style transformation and defunctionalization, or
from a structural operational semantics using the refocusing
construction of Danvy and
Nielsen~\cite{dvanhorn:Danvy-Nielsen:RS-04-26}.
